Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models

### 2020-06-15 analysis/paper/paper.Rmd
incorporated Sarah's comments from a Word version

---
title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
abstract: |
  Needs more...

  The QRF model presented here provided estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations. 
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. Recently released “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required. These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured. Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. For perhaps the first time, the necessity of tributary habitat restoration can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”

Pacific salmon (*Oncorhynchus spp.*) species have experienced large declines in abundance throughout much of their range [@Good2005]. Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity, both historic and contemporary, for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge. Reliable methods to better understand fish-habitat relationships as well as to estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity, and further, to better direct tributary rehabilitation efforts.

@Fausch1988 conducted a thorough review of attempts to predict the standing crop of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for some salmonid species. @Nickelson1992 found that juvenile coho were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale, however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship, which often results in weak predictive power.

A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships.

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density, neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves. However, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].

@@

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may only spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

1. Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
1. Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
1. Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
1. Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
1. Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

@@

Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, and single-pass electrofishing, as well as snorkeling. These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew / watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr, during summer months, using the natural log of fish densities as the response. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. However, the individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response, therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity, following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

@@

<!-- Using the QRF model, we predicted habitat capacity for juvenile rearing at all CHaMP sites within the interior Columbia River basin.  -->

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate, for use in predicting capacity within CHaMP watersheds, and one that did not, for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., fish/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots. These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another, so care should be taken when interpretting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

@@

Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)). QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the `r as.character(english(length(keep_wtsds)))` locations where comparisons were possible (Figure \@ref(fig:sr-figure)). Possible additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates. Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker). This favorable comparison provides strong validation, as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.

@@

## A Tool To Estimate Habitat Capacity

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, the models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.

@@

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity. However, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.

@@

Fish are mobile creatures, and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g. pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed. Therefore, we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population level scales. This is an efficient technique to leverage existing relationships for meaningful management decisions.

Our extrapolation model was focused on extrapolating to other master sample points, because that is the dataset available to us, but the methodology could be improved. Extrapolating to reaches on a stream network, as opposed to points on the landscape, could improve the interpretability of the results. This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used. Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and building a QRF model from that dataset. Even if the fish data was not done continuously, estimates of capacity could be made directly from the QRF model across the entire stream network, without the need for an extrapolation model.

>>>>

---
title: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
# csl: "../templates/canadian-journal-of-fisheries-and-aquatic-sciences.csl"
abstract: |
  Might still need some refining...

  Establishing metrics of quality habitat is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but rely on linear fish-habitat relationships and are limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout the Columbia River Basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile regression forest
highlights: |
  These are the highlights. 
---

@@

The decline of anadromous Pacific salmonids (*Oncorhynchus spp.*) across the Pacific Northwest, USA has prompted numerous actions aimed at reversing that trend. These actions are often categorized into four H’s – harvest modification, hatchery practices, hydro-system operations, and habitat rehabilitation. Problematically, there is substantial uncertainty regarding the degree of change that can be exerted across and within these categories, and what combination of changes will most cost-effectively and sustainably reduce mortality. The most recent “de-listing” criteria [@NOAA2016; @NOAA2016b] identified adult escapement targets at the population scale, providing a quantitative metric useful for evaluating the magnitude of survival improvements required. These abundance targets provide a benchmark against which habitat rehabilitation actions can be measured. Here we describe an approach for estimating life-stage specific habitat-based carrying capacity that can be used to quantitatively identify the magnitude of tributary habitat rehabilitation necessary to support de-listing. For perhaps the first time, the necessity of tributary habitat restoration can be demonstrated and the magnitude of required change can be placed in context with the other “H’s.”

Pacific salmon (*Oncorhynchus spp.*) species have experienced large declines in abundance throughout much of their range [@Good2005]. Declines can be partially attributed to lost or altered habitat, and thus, efforts to recover depleted salmon populations are replete with efforts to rehabilitate habitat used during the freshwater life-stages. Specifically, restoring salmonid carrying capacity through tributary rehabilitation actions has been identified as a key component of recovery efforts for salmon and steelhead (*Oncorhynchus mykiss*) in the Pacific Northwest, USA [@NOAA2016; @NOAA2016b]. Efforts have included increasing and improving existing habitat for both spawning adults and rearing juveniles. However, estimating habitat carrying capacity (both historic and contemporary) for various life-stages of Pacific salmon, as well as identifying important habitat characteristics that influence capacity, has been an ongoing but necessary challenge. Reliable methods to better understand fish-habitat relationships and estimate capacity are necessary to identify those salmon and steelhead life-stages that are limited by habitat capacity to better direct tributary rehabilitation efforts.

@Fausch1988 conducted a thorough review of attempts to predict the abundance of fish from measurable habitat covariates from 1950 to 1985, and found that the vast majority of multiple linear regression models failed to detect a significant fish-habitat signal. Since that review, there has been progress in identifying some fish-habitat relationships for some salmonid species. @Nickelson1992 demonstrated that juvenile coho were found in higher densities within pool habitat on the Oregon coast. Similarly, pool and pond densities were good predictors of coho smolt densities in western Washington [@Sharma2001]. @Bryant2009 found that juvenile coho abundance was positively related to large wood at the reach scale; however their results demonstrated a negative relationship between abundance and the number of pools. @Braun2011 similarly found positive associations between spawner densities of sockeye in the Fraser River and large wood, in addition to positive relationships to percent undercuts and percent pools. Densities of adult spawning coho were also higher in forested areas compared to urban or agricultural areas in the Snohomish River watershed [@Pess2002]. @Mossop2006 examined juvenile Chinook in the Yukon river and found positive correlations between the log of fish density and several metrics related to residual pool dimensions and large woody debris abundance as well as a negative correlation between fish density and gradient. These studies were focused on predicting observed fish densities, not necessarily capacity, and for most of them the predictive extent is limited to a particular watershed. In addition, they all assumed some form of linear fish-habitat relationship which often results in weak predictive power.

A number of studies have utilized other modeling approaches to elicit non-salmonid fish habitat relationships. @Dunham2002 used a quantile regression approach to show a negative relationship between cutthroat trout densities and the width:depth ratio of a stream for the upper quantiles of trout density. The same approach was also used to map the potential extent of sole in the English Channel and southern North Sea [@Eastwood2003]. Machine learning models such as boosted regression trees and random forests have been used to examine species biomass, diversity, and distribution for a number of different species [@Pittman2009; @Knudby2010; and @Compton2012]. @Wall2016 took a bioenergic approach to predict the capacity of a reach to support juvenile steelhead. The results from these studies highlight the importance and effectiveness of using techniques that can accommodate non-linear fish habitat relationships.

Most studies that have investigated fish-habitat relationships focus on predicting a species’ distribution (presence / absence) or the average abundance or density: neither of which can be easily manipulated to predict carrying capacity. Further, many of these studies focus on only one or two measures of habitat. @Sweka2010 estimated carrying capacity of Atlantic salmon parr using a quantile regression approach, but the only habitat covariate they included was cumulative drainage area. Traditionally, carrying capacity for salmonids has been estimated through stock-recruitment curves; however, this requires a long time-series of data with variety in the number of spawners which is not usually available [@Cramer2009].

@@

The fish abundance/density and habitat data used to fit the QRF model presented here were available from seven watersheds within the interior Columbia River basin, Pacific Northwest, USA. Within the interior Columbia River basin two major runs of Chinook salmon (*Oncorhynchus tshawytscha*) occur, stream-type (i.e., spring/summer run) and ocean-type (i.e., fall run), each characterized by different life history characteristics. Stream-type Chinook salmon enter freshwater earlier in the year, spawn in the upper reaches of a watershed, and the juveniles rear for up to 16 months in the freshwater before entering the ocean as smolts. Ocean-type Chinook salmon enter freshwater later (e.g. fall or winter) spawn lower in the watershed, and the juveniles may spend between several weeks and six months in freshwater before migrating to the ocean. Here we focus on juvenile stream-type Chinook, in particular the summer rearing period during low flow. Data presented here are from Chinook salmon populations in the Upper Columbia River spring-run and Snake River spring/summer-run Evolutionary Significant Units (ESU). The Upper Columbia spring-run ESU is listed as endangered under the Endangered Species Act, the Snake River spring/summer-run is listed as threatened [@NOAA2016; @NOAA2016b]. Hereafter, we refer to both ESUs simply as Chinook salmon.

@@

* Identify measured habitat characteristics that are most strongly associated with observed Chinook salmon parr abundance and density,
* Use paired fish and habitat measurements to elicit fish-habitat relationships for those habitat characteristics identified as important for determining fish abundance and density,
* Predict contemporary habitat carrying capacity at all sites where the important habitat characteristics are measured (i.e., CHaMP sites within the Columbia River basin),
* Extrapolate capacity predictions at CHaMP sites across a watershed using globally available attribute data to estimate the Chinook salmon parr capacity of that watershed, and
* Validate estimates of carrying capacity from our approach across multiple watersheds using independent estimates of capacity (e.g., spawner-recruit relationships).

@@

Juvenile fish surveys were conducted for Chinook salmon parr during the summer low-flow season at many of the same sites surveyed using the CHaMP protocol. Survey methods included mark-recapture, three-pass removal sampling, two-pass removal sampling, single-pass electrofishing, and snorkeling. These data were used to estimate Chinook salmon parr abundance at all CHaMP sites where fish survey data were available. Three-pass removal estimates used the Carle-Strub estimator [@Carle1978], following advice from @Hedger2013. Two-pass removal estimates used the estimator described by @Seber2002. Mark-recapture estimates used Chapman’s modified Lincoln-Peterson estimator [@Chapman1951] and were deemed valid if they met the criteria described in @Robson1964. These estimates were made using the *removal* function from the *FSA* package [@R-FSA] or the *closedp.bc* function from the *Rcapture* package [@R-Rcapture] in R software [@Rsoftware2019]. Snorkel counts were transformed to abundance estimates using paired snorkel-electrofishing sites to calibrate snorkel counts. For sites with invalid estimates or that were sampled with a single electrofishing pass, we developed an estimate of capture probability based on valid estimates, using a binomial generalized linear mixed effects model. Fixed effects were species, wetted width of the site, density of fish caught on the first pass and all possible two-way interactions. We included a random effect for fish crew / watershed. We used this model to predict abundances based on the number of fish caught on the first pass and any other covariates.

@@

Using the selected habitat covariates (Table \@ref(tab:select-mets-table)), we fit a QRF model to predict habitat rearing capacity for Chinook salmon parr during summer months using the natural log of fish densities as the response. After constructing a random forest, predictions of the mean response can be made by averaging the predictions of all trees, similar to the expected value predictions from a statistical regression model. The individual predictions from each tree, viewed collectively, describe the entire distribution of the predicted response; therefore, the random forest model can be used in the same way as other quantile regression methods to predict any quantile of the response. There were missing values for some habitat data; thus, any site visit with more than three missing covariates was removed from the dataset and the remaining missing habitat values were imputed using the *missForest* R package [@R-missForest; @Stekhoven2012]. We fit the QRF models using the *quantregForest* function from the *quantregForest* package [@R-quantregForest] in R software [@Rsoftware2019], incorporating data from `r nrow(qrf_data)` records (paired fish-habitat data) and `r as.character(english(nrow(sel_hab_mets)))` habitat covariates (`r round(nrow(qrf_data) / nrow(sel_hab_mets), 1)` data points per covariate) (Table \@ref(tab:select-mets-table)). The 90th quantile of the predicted distribution was used as a proxy for carrying capacity following the suggestion of @Sweka2010, and to avoid higher quantiles that draw from the very upper tails of observed fish density, where the variability of predictions may be influenced by small sample size issues.

@@

<!-- Drop the part about fitting an extrapolation model for areal density? -->

To predict capacity at larger spatial scales, such as the watershed or population, we developed two extrapolation models (linear and areal capacity) based on globally available attributes (GAA) which were available for the entirety of tributary habitat utilized by a given population. The GAA data used here was taken from the list of Generalized Random Tesselation Stratified master sample sites that the CHaMP sites were originally selected from [@Larsen2008; @Larsen2016]. Possible covariates included temperature range, growing degree days, an index of disturbance, the square root of cumulative drainage area, stream power, slope, channel type and watershed (Table \@ref(tab:gaa-covars)). The natural log of the CHaMP site capacity predictions was used as the response variable in a multiple linear regression model that incorporated the design weights of the CHaMP sites using the *svyglm* function from the *survey* package [@R-survey] in R software [@Rsoftware2019]. We fit two extrapolation models to each type of fish density, one that included watershed as covariate for use in predicting capacity within CHaMP watersheds, and one that did not for predicting everywhere else. We then made predictions of linear and areal capacity at all master sample sites throughout the interior Columbia River basin, generally spaced about one kilometer apart. Summaries of extrapolation model fit are shown in Table \@ref(tab:extrap-r2-table).

To summarize capacity at larger scales, the mean linear capacity (e.g., parr/m) of the master sample points along a particular tributary is multiplied by the length of that tributary. We first restricted the master sample points and lengths of streams to those with the domain of spring/summer Chinook salmon, as defined by StreamNet ([http://www.streamnet.org](http://www.streamnet.org) ) or using expert opinion from local biologists. The capacities of various tributaries could then be summed to estimate capacity at almost any spatial scale.

@@

A QRF model was fit using those metrics and the *quantregForest* package [@Meinshausen2006] in R [@Rsoftware2019] and the 90th quantile of the predicted distribution was used as a proxy for carrying capacity. After model fit, we examined the relative importance of each habitat covariate included in the model (Figure \@ref(fig:rel-imp-figure)). Moreover, QRF models allow one to visually examine the marginal effect of each habitat covariate on the quantile of interest using partial dependence plots. These plots show the marginal effect of changing a single habitat covariate while maintaining all other covariates at their mean values (Figure \@ref(fig:pdp-figure)). However, given that many habitat metrics are somewhat correlated, these marginal effects are often not independent of one another and care should be taken when interpreting them. After model fitting, the QRF model was used to predict habitat capacity at all CHaMP sites within the interior Columbia River basin.

@@

Estimates of Chinook salmon parr capacity from the QRF and extrapolation models were comparable to independent estimates from spawner-recruit data (Table \@ref(tab:sr-table), Figure \@ref(fig:sr-figure)). QRF estimates had overlapping confidence intervals with one or more of the Beverton-Holt, Ricker, or hockey stick model estimates in each of the `r as.character(english(length(keep_wtsds)))` locations where comparisons were possible (Figure \@ref(fig:sr-figure)). Possible additional uncertainty was not accounted for in estimates of spawners-per-redd or spawners-per-parr, which would increase the confidence intervals around spawner-recruit estimates and the overlap among estimates. Correlations between parr capacity estimates from the QRF model and spawner-recruit models ranged from 0.710 (Beverton-Holt) to 0.966 (Ricker). This favorable comparison provides strong validation as the spawner-recruit estimates of Chinook salmon parr capacity were developed from completely independent datasets and using entirely different methods.

@@

## A Tool to Estimate Habitat Capacity

In this study, we developed a novel approach to estimate the capacity of habitat to support Chinook salmon parr during summer months in wadeable streams. Our model can be used to quantify juvenile rearing capacity in Chinook salmon watersheds or populations and, in turn, quantify the magnitude of tributary habitat rehabilitation necessary to support Endangered Species Act delisting. The QRF and extrapolation models presented here provide useful tools towards the prioritization, implementation, and evaluation of habitat rehabilitation actions to recover depleted salmon populations. Moreover, these models can be applied to multiple stages within the life cycle (e.g., parr, smolt, adult). Estimates of habitat carrying capacity for multiple life stages will allow biologists and managers to identify what life-stage specific habitat patches may be limiting. As an example, QRF models and associated extrapolation models may demonstrate that habitat for a given population is sufficient to support adult spawning required to achieve delisting targets, but that juvenile rearing capacity may not be sufficient to support the target abundance. In such a case, habitat rehabilitation actions may be most cost-effectively and sustainably directed towards improving juvenile rearing habitat. Models to estimate habitat carrying capacity for multiple life stages will help to better direct habitat restoration actions.

@@

The biggest driver of capacity identified in this study is stream size, whether measured by wetted width or discharge, which should be unsurprising since we are using fish per meter as our response. In many ways, these metrics define habitat quantity; however, other metrics used in our QRF model help define habitat quality, such as cooler temperatures in August, less pool-tail fine sediment, and higher channel unit frequencies (a measure of habitat complexity and another way of saying more pool-riffle interfaces) and fish cover.

@@

Fish are mobile creatures and determining the appropriate spatial scale to estimate how their capacity may be determined by habitat characteristics is important. In the summer, for Chinook parr, there is clearly movement between multiple channel units (e.g. pool, riffle, run), suggesting that fish are utilizing habitat at a larger scale than the channel unit. However, it is unlikely that they are moving up and down the entire watershed and we believe the 200 - 500 m reaches used in this study are an appropriate scale to capture the fish-habitat relationships that define carrying capacity. At the same time, we acknowledge that managers, life-cycle modelers, and others are often interested in capacity estimates at larger spatial scales. While our QRF model can provide site-specific estimates of carrying capacity derived from paired fish-habitat data, our extrapolation model allows for estimates at larger spatial extents, such as watershed and population level scales. This is an efficient technique to leverage existing relationships for meaningful management decisions.

Our extrapolation model was focused on extrapolating to other master sample points, because that is the dataset available to us, but the methodology could be improved. Extrapolating to reaches on a stream network, as opposed to points on the landscape, could improve the interpretability of the results. This would require a stream network with relevant attributes attached to each reach, similar to the GAAs we used. Another approach could be to move towards sampling habitat in a more spatially continuous fashion, covering most or all of a watershed, and building a QRF model from that dataset. Even if the fish data were not collected continuously, estimates of capacity could be made directly from the QRF model across the entire stream network without the need for an extrapolation model.
